---
title: "Peer Assessment 1"
author: "Charles Chisanga"
date: "18th December 2015"
---
###About
This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

##Purpose
The purpose of this project was to practice:

* loading and preprocessing data
* imputing missing values
* interpreting data to answer research questions

## Data
The data for this assignment can be downloaded from the course web site:
Dataset: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip [52K]
The variables included in this dataset are:
steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)
date: The date on which the measurement was taken in YYYY-MM-DD format
interval: Identifier for the 5-minute interval in which measurement was taken
The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

## Loading and preprocessing the data

Download, unzip and load the data into data frame "data". 
```{r}
if(!file.exists("getdata-projectfiles-UCI HAR Dataset.zip")) {
        rrdata <- tempfile()
        download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",rrdata)
        unzip(rrdata)
        unlink(rrdata)
}

data <- read.csv("activity.csv")
```

## What is mean total number of steps taken per day?
Total number of steps/days, Make a histogram, Calculate and report mean and median of total number of steps/day
```{r} 
steps_per_day <- aggregate(steps ~ date, data, sum)
hist(steps_per_day$steps, main = paste("Total Steps per Day"), col="blue", xlab="Number of Steps")
stepmean <- mean(steps_per_day$steps)
stepmean
stepmedian <- median(steps_per_day$steps)
stepmedian
```

The `mean` is `r stepmean` and the `median` is `r stepmedian`.

## What is the average daily activity pattern?

* Calculate average steps for each interval for all days. 
* Plot the Average Number Steps per Day by Interval. 
* Find interval with most average steps. 
```{r}
steps_per_interval <- aggregate(steps ~ interval, data, mean)

plot(steps_per_interval$interval,steps_per_interval$steps, type="l", xlab="Interval", ylab="Number of Steps", main="Average Number of Steps per Day by Interval")

max_interval <- steps_per_interval[which.max(steps_per_interval$steps),1]
```

The 5-minute interval, on average across all the days in the data set, contains the maximum number of steps is `r max_interval`.

## Impute missing values. Compare imputed to non-imputed data.
Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

Missing values were imputed by inserting the average for each interval. Thus, if interval 10 was missing on 10-02-2012, the average for that interval for all days (0.1320755), replaced the NA. 
```{r}
missing_values <- sum(!complete.cases(data))
imputed_data <- transform(data, steps = ifelse(is.na(data$steps), steps_per_interval$steps[match(data$interval, steps_per_interval$interval)], data$steps))
```

Zeroes were imputed for 10-01-2012 because it was the first day and would have been over 9,000 steps higher than the following day, which had only 126 steps. NAs then were assumed to be zeros to fit the rising trend of the data. 
```{r}
imputed_data[as.character(imputed_data$date) == "2012-10-01", 1] <- 0
```

Recount total steps per day and create Histogram. 
```{r}
steps_per_day_i <- aggregate(steps ~ date, imputed_data, sum)
hist(steps_per_day_i$steps, main = paste("Total Steps per Day"), col="blue", xlab="Number of Steps")

#Create Histogram to show the difference. 
hist(steps_per_day$steps, main = paste("Total Steps per Day"), col="brown", xlab="Number of Steps", add=T)
legend("topleft", c("Imputed", "Non-imputed"), col=c("blue", "brown"), lwd=10)
```

Calculate new mean and median for imputed data. 
```{r}
stepmean.i <- mean(steps_per_day_i$steps)
stepmean.i
stepmedian.i <- median(steps_per_day_i$steps)
stepmedian.i
```


Calculate difference between imputed and non-imputed data.
```{r}
mean_diff <- stepmean.i - stepmean
mean_diff
med_diff <- stepmedian.i - stepmedian
med_diff
```

Calculate total difference.
```{r}
total_diff <- sum(steps_per_day_i$steps) - sum(steps_per_day$steps)
total_diff
```

* The imputed data mean is `r stepmean.i`
* The imputed data median is `r stepmedian.i`
* The difference between the non-imputed mean and imputed mean is `r mean_diff`
* The difference between the non-imputed mean and imputed mean is `r med_diff`
* The difference between total number of steps between imputed and non-imputed data is `r total_diff`. Thus, there were `r total_diff` more steps in the imputed data.


## Are there differences in activity patterns between weekdays and weekends?
Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

``` {r}
weekdays <- c("Monday", "Tuesday", "Wednesday", "Thursday", 
              "Friday")
imputed_data$dow = as.factor(ifelse(is.element(weekdays(as.Date(imputed_data$date)),weekdays), "Weekday", "Weekend"))

steps_per_interval_i <- aggregate(steps ~ interval + dow, imputed_data, mean)
```

Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.

```{r}
library(lattice)
xyplot(steps_per_interval_i$steps ~ steps_per_interval_i$interval|steps_per_interval_i$dow, main="Average Steps per Day by Interval",xlab="Interval", ylab="Steps",layout=c(1,2), type="l")
```
